---
title: "Environmental outlier detection with bootstrapping and principle component analysis."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental outlier detection with bootstrapping and principle component analysis.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"#, error = TRUE
)
```

```{r libraries, echo=TRUE, warning=FALSE, message=FALSE}

library(specleanr)

```
**Environmental outlier check for fish species from the Danube River Basin** 

The workflow for environmental outlier detection and removal is similar across taxa, regions, or ecological realms. However, we included the **`check_names()`** function to cater for fish species names exhaustively. In this worked example, we tried the functionalities on the fish species from the Danube River Basin, with extracts of species records from Joint Danube Survey (JDS) and EFI+ data archived in the package. We complimented the data with fish species occurrences from online sources including Global Biodiversity Information Facility (GBIF), iNaturalist, and VertNET using the **`getdata()`** function. They are basically five steps, including: 1) Data acquisition and harmonization; 2) Precleaning and predictor extraction 3) outlier detection 4) identification of clean data and suitable method 5) developing species distribution models (optional).


1. **Data acquisition: a) Collate species species records: offline and online data**

The species records were obtained from the archived datasets extracted from the Joint Danube Survey (https://www.danubesurvey.org/jds4/) and EFIPLUS (Logez et al., 2012).  To compliment species records, we used the **`getdata()`** function to retrieve data from the GBIF (https://www.gbif.org/), VertNet (http://www.vertnet.org/) and iNaturalist ( https://www.inaturalist.org/) for _Squalius cephalus_, _Salmo trutta_, _Thymallus thymallus_, and _Anguilla anguilla_. For online data, we limited data to 50 records from each data source to reduce on the execution time.

**NOTE**

This workflow may fail if the particular settings such as blocking of IP address to GIBF, iNaturalist, FishBase, and VertNet may prevent user from accessing the online data. Since, this may user-to-user basis, it is beyond the scope of this package to address such errors.

```{r datasoruces, warning=FALSE}
#==========================
#Step 1ai. Obtain Local data sources (archived in this package)
#=========================

data(efidata) #Data extract from EFIPLUS data

data(jdsdata) #Data extract from JDS4 data 

#===================================
#Step 1aii: Retrieve online data for the species: polygon to limit the extent to get records.
#=====================================
danube <- sf::st_read(system.file('extdata', "danube.shp.zip",
                                  package = 'specleanr'), quiet=TRUE)


df_online <- getdata(data = c("Squalius cephalus", 'Salmo trutta', 
                              "Thymallus thymallus","Anguilla anguilla"), 
                                extent = danube,
                                gbiflim = 50, 
                                inatlim = 50, 
                                vertlim = 50, 
                     verbose = F)

dim(df_online)

```

**Merging and harmonizing species records from different sources**

The online data sources from **`getdata()`** and local files are merged using the **`match_datasets()`** function. Five columns are harmonized while combining data from different sources: the country, species names, latitude/longitude columns, and dates. The Darwin Core standard names are country, species, decimalLatitude, decimalLongitude, and dates (Wieczorek et al., 2012). So, if the local dataset has a different name for standard names, the user should indicate it. For example, in JDS data, the species column is labeled **speciesname**, shown in the species parameter for automatic renaming and merging with other datasets. 
* Note: The user should indicate all dataset names in the list.
* `check_names()` is used to clean species names in terms of synonyms or spellings, based on FishBase (https://www.fishbase.se/). This function generates another column **speciescheck** that contain the clean names.

```{r merging and harmonising species records handling, warning=FALSE}

mergealldfs <- match_datasets(datasets = list(efi= efidata, jds = jdsdata, 
                                        onlinedata = df_online),
                country = c('JDS4_sampling_ID'),
                lats = 'lat', lons = 'lon',
                species = c('speciesname', 'scientificName'))

#Species names are re-cleaned since the species names from vertnet are changed.

cleannames_df <- check_names(data = mergealldfs, colsp = 'species', pct = 90, 
                             merge = TRUE, verbose = TRUE)

#Filter out species from clean names df where the species names such as synonyms like Salmo trutta fario chnaged to Slamo trutta

speciesfiltered <- cleannames_df[cleannames_df$speciescheck %in%
                                   c("Squalius cephalus", 'Salmo trutta', 
                                     "Thymallus thymallus","Anguilla anguilla"),]

```

1. **Data acquisition: b) Environmental predictors**

We used WORLDCLIM data archived in the package to enable users to test the package functions seamlessly. For direct interaction with the WORDCLIM data, please visit (https://www.worldclim.org/) and the **`geodata`** package for download in user-customized workflows. WORLDCLIM data has 19 bioclimatic variables (https://www.worldclim.org/data/bioclim.html), including;

* `BIO1` = Annual Mean Temperature
* `BIO2` = Mean Diurnal Range (Mean of monthly (max temp - min temp))
* `BIO3` = Isothermality (BIO2/BIO7) (×100)
* `BIO4` = Temperature Seasonality (standard deviation ×100)
* `BIO5` = Max Temperature of Warmest Month
* `BIO6` = Min Temperature of Coldest Month
* `BIO7` = Temperature Annual Range (BIO5-BIO6)
* `BIO8` = Mean Temperature of Wettest Quarter
* `BIO9` = Mean Temperature of Driest Quarter
* `BIO10` = Mean Temperature of Warmest Quarter
* `BIO11` = Mean Temperature of Coldest Quarter
* `BIO12` = Annual Precipitation
* `BIO13` = Precipitation of Wettest Month
* `BIO14` = Precipitation of Driest Month
* `BIO15` = Precipitation Seasonality (Coefficient of Variation)
* `BIO16` = Precipitation of Wettest Quarter
* `BIO17` = Precipitation of Driest Quarter
* `BIO18` = Precipitation of Warmest Quarter
* `BIO19` = Precipitation of Coldest Quarter

```{r environmental parameters from WORLDCLIM}

#Get climatic variables from the package folder

worldclim <- terra::rast(system.file('extdata/worldclim.tiff', package = 'specleanr'))

```

2. **Precleaning and environmental data extraction**

Here, 
* The duplicate records are removed if points they are obtained from the same location for the same species. 
* The missing values coordinates are removed. 
* The environmental predictors are extracted from the raster layers (WORLDCLIM).
* The user can set the minimum point for the species to be retianed for further analyis.
* The bounding box can be set to limit the extent of data extraction. For this case, we used the basin layer for the Danube Basin was obtained from Hydrography90m (https://hydrography.org/hydrography90m/hydrography90m_layers).
* The user can either return a dataframe or list of the cleaned data. Important in the next steps.

```{r precleanand, echo=TRUE}

#Get basin shapefile to delineate the study region: optional

danube <- sf::st_read(system.file('extdata', 'danube.shp.zip', 
                                  package = 'specleanr'), quiet=TRUE)

#For multiple species indicate multiple TRUE
multipreclened <-  pred_extract(data= speciesfiltered, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'speciescheck',
                             bbox  = danube,  
                             list= TRUE, 
                             minpts = 10, merge = FALSE)
names(multipreclened)


ttdata <- speciesfiltered[speciesfiltered[,'speciescheck'] %in%c("Thymallus thymallus"),]
dim(ttdata)

#For single species indicate multiple to FALSE
ttrefdata<-  pred_extract(data= ttdata, raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'speciescheck',
                             bbox  = danube,
                             list= TRUE, 
                             minpts = 10)
dim(ttrefdata)

```

3. **Outlier detection**
Multiple outlier detection are set. This package contains 20 outlier detection methods and the user can run **`extractMethods()`** to get the allowed methods. They are categorized into univariate, multivariate and species ecological ranges.
* `var` is the predictor to be used univariate methods.
* `exclude` allows to remove predictors that user deems unnecessary. For example, the coordinates, since the multivariate methods consider the whole dataset.

**No bootstrapping and principle component analysis**

```{r outlierdetection, echo=TRUE, message=FALSE, warning=FALSE}

#For multiple species: default settings

multioutliers <- multidetect(data = multipreclened,
                      multiple = TRUE,
                      var = 'bio6',
                     exclude = c('x','y'),
                      methods = c('zscore', 'adjbox',
                                                'logboxplot', 'distboxplot',
                                                'iqr', 'semiqr',
                                                'hampel','kmeans',
                                                'jknife', 'onesvm',
                                                'iforest'))
#single species:default settings

ttoutlier<- multidetect(data = ttrefdata,
                      multiple = FALSE,
                      var = 'bio6',
                      output = 'outlier',
                      exclude = c('x','y'),
                      methods = c('zscore', 'adjbox',
                                  'logboxplot', 'distboxplot',
                                  'iqr', 'semiqr',
                                  'hampel','kmeans',
                                  'jknife', 'onesvm',
                                  'iforest'))

```


4. **Outlier visualization**

* `ggoutliers` are based in ggplot2, so it can be modified based on user needs. x: is the output for outlier detection, y is the species name or index for multiple species, and raw = TRUE if the number of outliers are the displayed, otherwise the proportion of outliers to the total number of records will be plotted.

```{r visualisation, warning=FALSE, fig.width = 6, fig.height= 5, fig.align='center'}
#for multiple species
ggoutliers(multioutliers)

#for single species
ggoutliers(ttoutlier)

```

**Identify the best threshold using loess model.**

The local regression is used to optimize and identify the best threshold for denoting the point as an absolute outlier. We fit the local region model between the data retained at every threshold, and we identify a maxima when the number of records retain are number of records retained does not significantly vary with an increased increase in the threshold.

```{r threshold identifcation, fig.width = 6, fig.height= 5, fig.align='center'}

ttoptimal <- optimal_threshold(refdata = ttrefdata, outliers = ttoutlier, plot = TRUE)

#obtain the optimal thresholds for multiple species 
#plotting is not allowed for multiple species.

multioptimal <- optimal_threshold(refdata = multipreclened, outliers = multioutliers)

```


5. **Extracting clean data from the reference data (precleaned data in step 2).**

The user sets a threshold ranging from 0.1 to 1 but its advisable to set a value above 0.5 to include above 50% of the methods. **threshold** is the value indicating the proportion of methods to be used to classify a record as a true outlier. For example, a threshold of 0.6 means that at least in the 4 of the 6 methods noted during outlier detection in step 3. We used the loess method for identifying the optimal threshold.

```{r extract clean dataset}

multspecies_clean <- extract_clean_data(refdata = multipreclened, 
                                        outliers = multioutliers, 
                                        loess =  TRUE)
head(multspecies_clean)

ttclean<- extract_clean_data(refdata = ttrefdata, outliers = ttoutlier, loess = TRUE)

head(ttclean)

labeldata_mult <- classify_data(refdata = multipreclened, 
                                outliers = multioutliers, 
                                EIF = TRUE)
head(labeldata_mult)


labeldata_tt <- classify_data(refdata = ttrefdata, outliers = ttoutlier, EIF = TRUE)
head(labeldata_tt)

```

6. **Visualize labeled data in environmental space.**

```{r 2d plots, fig.width = 6, fig.height= 5, fig.align='center'}

#multiple species 

ggenvironmentalspace(qcdata = labeldata_mult, 
                     xvar = 'bio1', 
                     yvar = "bio18", 
                     xlab = "Annual mean temperature",
                     ylab = "Precipitation of Warmest Quarter",
                     scalecolor = 'viridis',
                     ncol = 2, 
                     nrow = 2,
                     pointsize = 2)

#for single species
ggenvironmentalspace(qcdata = labeldata_tt, 
                     xvar = 'bio1', 
                     yvar = "bio18",
                     xlab = "Annual mean temperature",
                     ylab = "Precipitation of Warmest Quarter",
                     scalecolor = 'viridis',
                     pointsize = 2)

```

### Using bootstrapping during outlier detection

* The number of records **`maxrecords´** in reference dataset for _Thymallus thymallus_ is 99. Therefore, to implement bootstrapping, indicate the maximum higher than the nrows in reference dataset otherwise bootstrap will not be implemented.
* The number of bootstraps, **`nb´** are user-defined.

```{r bootstrappingoutlier detection}

ttoutlierboot <- multidetect(data = ttrefdata,
                      multiple = FALSE,
                      var = 'bio6',
                      exclude = c('x','y'),
                      methods = c('zscore', 'adjbox',
                                  'logboxplot', 'distboxplot',
                                  'iqr', 'semiqr',
                                  'hampel','kmeans',
                                  'jknife', 'onesvm',
                                  'iforest'),
                      bootSettings = list(run = TRUE, maxrecords = 100, nb = 10))

```
### Visulise outliers after bootstrapping 

```{r visualisationboot, fig.align='center', fig.height=5, fig.width=6, warning=FALSE}

ggoutliers(ttoutlierboot)

```

### Classify data to obtain labels 

```{r classifyboot, warning=FALSE, fig.width = 6, fig.height= 5, fig.align='center'}

labeldata_boot <- classify_data(refdata = ttrefdata, outliers = ttoutlierboot)

```

### Visualise after bootstrapping 

```{r ggspaceboot, warning=FALSE, fig.width = 6, fig.height= 5, fig.align='center'}

ggenvironmentalspace(qcdata = labeldata_boot, 
                     xvar = 'bio1', 
                     yvar = "bio18",
                     xlab = "Annual mean temperature",
                     ylab = "Precipitation of Warmest Quarter",
                     scalecolor = 'viridis',
                     pointsize = 2)

```

**Notes**

When bootstrapping is applied, the very strong outlier was changed to moderate label


**References**
1. Wieczorek, J., Bloom, D., Guralnick, R., Blum, S., Döring, M., Giovanni, R., Robertson, T., & Vieglais, D. (2012). Darwin core: An evolving community-developed biodiversity data standard. PLoS ONE, 7(1). https://doi.org/10.1371/journal.pone.0029715

