---
title: "Environmental outlier detection in fish species records."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental outlier detection in fish species records.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"#, error = TRUE
)
```

```{r libraries, echo=TRUE, warning=FALSE, message=FALSE}

library(specleanr)

```
**This workflow** 

1. **Collate species and environmental datasets**

```{r datasoruces, warning=FALSE}

#1a. Get basin shapefile to delineate the study region: optional

danube <- system.file('extdata/danube/basinfinal.shp', package = 'specleanr')

danube_basin  <-     sf::st_read(danube, quiet=TRUE)

#==========================
#Step 1b. Obtain Local data sources (Optional if not available)
#=========================

data(efidata) #Data extract from EFIPLUS data

data(jdsdata) #Data extract from JDS4 data 

#==========
#Step 1c. Download environmental data

worldclim <- system.file('extdata/worldclim.tiff', package = 'specleanr')
worldclim <- terra::rast(worldclim)

```


2. **Harmonizing data and collating species ecological parameters and nomenclature**
#Step 3. Merge local data files 
#Will also enable to generate the species list of interest
#The efiplus latitude and longitude are in the standard format names (deimalLatitude and decimalLongitude), so they will not be changed.

```{r datahandling}

localdf <- match_datasets(datasets = list(jds=jdsdata, efi=efidata),
                    lats = 'lat', lons = 'lon', 
                    species = c('speciesname','scientificName'),
                    country = c('JDS4_site_ID'),
                    date = c('sampling_date', 'Date'))


#Step 4. Harmonize all the species names by crosschecking with FishBase
#Keep merge = TRUE to stick a cleaned species names column
#NAs species will be automatically dropped

localdf_check <- check_names(localdf, colsp = 'species',verbose = FALSE, pct = 90, 
                             merge = TRUE)

#==============

#Retrieve online data from Global Information Facility, VertNET and iNaturalist
#Note: Using the cleaned species column species
#If local data sources are NULL or optional, this will be mandatory and vice versa

lagoco_online <- getdata(data = c('Lampetra planeri', 'Gobio gobio','Cottus gobio'), 
                                bbox = danube_basin,
                                gbiflim = 10, 
                                inatlim = 10, 
                                vertlim = 10, verbose = F)

```

3. **Further data handling**

```{r mergeclean}

#======================
#Already online and local data sources prepared

#1. Online parameter only accepts dataonline class
#2. The offline lists must be be named
#======

onlinedf <- extract_online(online = lagoco_online)

#match all data sets

matchalldf <- match_datasets(datasets = list(efi= efidata, jds = jdsdata, 
                                        onlinedata = onlinedf),
                country = c('JDS4_sampling_ID'),
                lats = 'lat',
                lons = 'lon',
                date = c('sampling_date', 'Date', 'dates', 'date'),
                species = c('speciesname', 'scientificName'))

#Species names are re-cleaned since the species names from vertnet are changed.

cleandata <- check_names(data = matchalldf, colsp = 'species', pct = 70, merge = TRUE)

```

4. **Precleaning and environmental data extraction**

```{r precleanand, echo=TRUE}

#two species

#Generate lists of species data

precleaned <-  pred_extract(data= cleandata, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'speciescheck',
                             bbox  = danube_basin,  
                             multiple = TRUE, 
                             list= TRUE, 
                             minpts = 10, merge = FALSE)

salmo1 <- cleandata[cleandata[,'speciescheck'] == "Salmo trutta", ]

precleaned1 <-  pred_extract(data= salmo1, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'speciescheck',
                             bbox  = danube_basin,  
                             multiple = FALSE, 
                             list= TRUE, 
                             minpts = 10, merge = FALSE)

```




```{r outlierdetection, echo=TRUE, message=FALSE, warning=FALSE}

#For outliers, the output is set to outlier

out_df <- multidetect(data = precleaned,
                      multiple = TRUE,
                      var = 'bio6',
                      output = 'outlier',
                     exclude = c('x','y'),
                      methods = c('zscore', 'adjbox','iqr', 'semiqr','hampel','kmeans'))
#single species
out_df1 <- multidetect(data = precleaned1,
                      multiple = FALSE,
                      var = 'bio6',
                      output = 'outlier',
                      exclude = c('x','y'),
                      methods = c('zscore', 'adjbox','iqr', 'semiqr','hampel','kmeans'))

```


6. **Data analysis and visualization**

```{r visualisation, warning=FALSE, fig.width = 6, fig.height= 5, fig.align='center'}

plot(out_df, y=8, raw=TRUE)

```

```{r extract clean dataset}

cleanfishdata <- clean_data_extract(refdata = precleaned, outliers = out_df, threshold = 0.6)

cleanfishdata1 <- clean_data_extract(refdata = precleaned1, outliers = out_df1, threshold = 0.6)

```

```{r model comparison}

modelcompare <- modelcomparison(refdata = precleaned1, outliers = out_df1, raster = worldclim, 
                                lat = 'y', lon = 'x', full = FALSE, metrics = 'dep')

mp <- get_performance(modeloutput = modelcompare)

```

