---
title: "General outlier detection for univariate datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General outlier detection for univariate datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(specleanr)
```

### Introduction to general data outlier detection 

* The presence of outliers is a general concern even is data which is not specifically for species distribution modelling. Also, the same approach of adhoc selection of outlier detection method is considered to detect and remove outliers in the data, which increased subjectivity. Therefore, we extend *specleanr* principle of ensembling multiple outlier detection methods to identify *absolute outliers* in the datasets which can later be removed. 

* The same process is followed but the no data extraction and evaluating model performance is required.

* Below is a detailed workflow for objectively detecting and removing outlier in *iris* dataset which is incorporated in the *dataset* package in *R programming Language*

```{r Getting iris and adding some outliers}

irisdata1 <- iris

#introduce outlier data and NAs

rowsOutNA1 <- data.frame(x= c(344, NA,NA, NA),
                         x2 = c(34, 45, 544, NA), 
                         x3= c(584, 5, 554, NA),
                         x4 = c(575, 4554,474, NA), 
                         x5 =c('setosa', 'setosa', 'setosa', "setosa"))

colnames(rowsOutNA1) <- colnames(irisdata1)


dfinal <- rbind(irisdata1, rowsOutNA1)

```

### Detecting outlier in changed iris dataset*

We can only use univariate methods to detect only in in variable such as Sepal.Length or we can exclude the species column and also use multivariate methods such as isolation forest, Mahalanobis outlier detection method or One class support vector machines. To identify the methods allowed in this package, run `**extractMethod()**`


**NOTE**
* Because we are considering univariate analysis, the parameter **sdm** is set to **FALSE**.

* Also, all multivariate outlier detection methods are not necessary for univariate datasets. The function **extractMethods()** can be used to identify the different methods allowed.

* Indicate **na.inform** to show how NAs are handled in the dataset. If percentage NAs in a column are greater than then **missingness** parameter, then that particular column will be removed. Otherwise, the rows with NAs will be removed using **na.omit** to avoid failing the outlier detection methods. In summary, increasing **missingness** may lead to loss many rows especially if any column has more missing values.

**Filter out only setosa data before outlier detection**

```{r outlier detection using iris dataset}

setosadf <- dfinal[dfinal$Species%in%"setosa",c("Sepal.Width", 'Species')]

setosa_outlier_detection <- multidetect(data = setosadf, 
                                var = 'Sepal.Width', 
                                multiple = FALSE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = FALSE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)

#extractMethods()

```

### Visualiss the number of outliers detected by each method 

```{r ploting, fig.width = 6, fig.height= 4, fig.align='center' }

ggoutliers(setosa_outlier_detection)


```

### Obtaining quality controlled dataset using loess method or data labelling 

```{r clean data extraction and labelling}

setosa_qc_loess <- extract_clean_data(refdata = setosadf, 
                                      outliers = setosa_outlier_detection, loess = TRUE)

#clean dataset
nrow(setosa_qc_loess)

#reference data
nrow(setosadf)

setosa_qc_labeled <- classify_data(refdata = setosadf, outliers = setosa_outlier_detection)

```

### Visualise labelled quality controlled dataset

```{r clean data extraction and labelling_multiple, fig.height=4.1, fig.width=4.5, fig.align='center'}


ggenvironmentalspace(setosa_qc_labeled, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")

```

### For multiple species

**NOTE**

* For multiple groups, the parameter **multiple** is changed to **TRUE** and the **var_col** should be provided as demonstrated below.

```{r outlier detection using multiple}

multspp_outlier_detection <- multidetect(data = dfinal, 
                                var = 'Sepal.Width', 
                                multiple = TRUE,
                                var_col = "Species",
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = FALSE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)

```

### Visualise the number of outliers detected by each method 

```{r ploting_single, fig.width = 6, fig.height= 4, fig.align='center' }

ggoutliers(multspp_outlier_detection)


```

### Obtaining quality controlled dataset using loess method or data labelling 

```{r clean data extraction and labelling mult}

multsp_qc_loess <- extract_clean_data(refdata = dfinal, 
                                      outliers = multspp_outlier_detection,
                                      var_col = 'Species',
                                      loess = TRUE)

#clean dataset
nrow(multsp_qc_loess)

#reference data
nrow(dfinal)

multi_qc_labeled <- classify_data(refdata = dfinal, 
                                      outliers = multspp_outlier_detection,
                                  var_col = 'Species')

```

### Visualise labelled quality controlled dataset

```{r visualise data labelling, fig.height=4.5, fig.width=6.5, fig.align='center'}

ggenvironmentalspace(multi_qc_labeled, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")

```


*The package is undergoing peer review for publication*








