---
title: "General outlier detection and removal using ensembling principle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General outlier detection and removal using ensembling principle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(specleanr)
```

The presence of outliers is a general concern even is data which is not specifically for species distribution modelling. Also, the same approach of adhoc selection of outlier detection method is considered to detect and remove outliers in the data, which increased subjectivity. Therefore, we extend *specleanr* principle of ensembling multiple outlier detection methods to identify *absolute outliers* in the datasets which can later be removed. 
* The same process is followed but the no data extraction and evaluating model performance is required.
* Below is a detailed workflow for objectively detecting and removing outlier in *iris* dataset which is incorporated in the *dataset* package in *R programming Language*

```{r Getting iris and adding some outliers}


irisdata1 <- iris
#add outlier data and NAs
rowsOutNA1 <- data.frame(x= c(344, NA,NA, NA), 
                         x2 = c(34, 45, 544, NA), 
                         x3= c(584, 5, 554, NA),
                        x4 = c(575, 4554,474, NA), 
                        x5 =c('Setosa', 'Setosa', 'Setosa', "Setosa"))

colnames(rowsOutNA1) <- colnames(irisdata1)


dfinal <- rbind(irisdata1, rowsOutNA1)


```

*Detecting outlier in changed iris dataset*
We can only use univariate methods to detect only in in variable such as Sepal.Length or we can exclude the species column and also use multivariate methods such as isolation forest, Mahalanobis outlier detection method or One class support vector machines. To identify the methods allowed in this package, run `**extractMethod()**`

```{r}
outlieriris_univ <- multidetect(data = dfinal, var = 'Sepal.Length', multiple = FALSE,
                       methods = c('mixediqr', 'logboxplot','iqr', 'distboxplot',
                                   'jknife', 'semiqr', 'hampel'),
                       silence_true_errors = FALSE, missingness = 0.1)

outlieriris_mult <- multidetect(data = dfinal, var = 'Sepal.Length', multiple = FALSE,
                       exclude = "Species",
                       methods = c('mixediqr', 'logboxplot','iqr', 'distboxplot', 
                                   'jknife', 'semiqr','hampel', 'iforest', 
                                   'lof', 'mahal'),
                       silence_true_errors = TRUE, missingness = 0.1)

```

*Plotting graphs to identify the different outliers detected by each method*

```{r ploting, fig.width = 6, fig.height= 4, fig.align='center' }

#Only univariate methods

ggoutliers(outlieriris_univ)

#multivariate methods considered

ggoutliers(outlieriris_mult)

```

*clean data extraction* 

```{r clean data extraction}

cleandata1 <- extract_clean_data(refdata = dfinal, outliers = outlieriris_univ, loess = TRUE)

#clean dataset
nrow(cleandata1)

#reference data
nrow(dfinal)


cleandata2 <- extract_clean_data(refdata = dfinal, outliers = outlieriris_mult, loess = TRUE)

#clean dataset
nrow(cleandata2)

#reference data
nrow(dfinal)

```


## Reference

1. Ab under review






