---
title: "Detecting environmental outliers in species distribution models for plants."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting environmental outliers in species distribution models for plants.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}

library(specleanr)

```

***Detect environmental outliers for species occurrences in *Fagus sylvatica*, and *Populus nigra* records based on bioclimatic data***.


To test the workflow on two plant species, namely the black poplar (*Populus nigra*) , which is a pioneer wind-pollinated deciduous tree from the Salicaceae family (de Rigo et al., 2016). The tree is distributed across Europe, Asia, and northern Africa (de Rigo et al., 2016; Harvey-Brown et al., 2017) and predominantly in flood plains in mixed forests (Harvey-Brown et al., 2017). The mature tree lives up to 300 to 400 years (de Rigo et al., 2016), and it has been assessed as data deficient according to IUCN RedList (Harvey-Brown et al., 2017). Secondly, we considered the European beech or common beech (*Fagus sylvatica*), which is also a deciduous hardy tree species that can tolerate various soil habitats but is not tolerant of waterlogged areas (Barstow et al., 2018). Although mature individuals have declined, the tree species IUCN RedList threat status is assessed as Least Concern (Barstow et al., 2018).

1. **Data acquisition: a) species records**

Data was retrieved from the iNaturalist (https://www.inaturalist.org/) and Global Biodiversity Information Facility-GBIF (https://www.gbif.org/). We use the **`getdata()`** function, built around rvertnet, rgbif, and rinat packages. We limited the occurrence records from GBIF to 700 and iNaturalist to 100. 
* **Note**: If the user has locally saved data, then after extracting the online data, the **`match_datasets()`** function can be used to merge all the data sets. 

* We set the bounding box **`(extent)`** parameter for **`getdata()`** to return only records within the Danube basin. However, the user can set this to suit the area of concern. If not set, the user data will download records globally, which is time-consuming and likely to break during the download. Therefore, it is strongly advisable to set the **`bounding box`**. Also, if a shapefile is available, the user can provide it instead of the bounding box, and the function will automatically extract the bounding box.

```{r get species occurences}

plantdf <- getdata(data = c( "Populus nigra", "Fagus sylvatica"), 
                    gbiflim = 700, inatlim = 100,
                     hasCoordinate = TRUE, 
                   extent = list(xmin = 8.15250, ymin = 42.08333, xmax=29.73583, ymax = 50.24500),
                   verbose = FALSE, warn = FALSE)


```
1. **Data acquisition: b) Environmental predictors**

We used WORLDCLIM data archived in the package to enable users to test the package functions seamlessly. For direct interaction with the WORDCLIM data, please visit (https://www.worldclim.org/) and the **`geodata`** package for download in user-customized workflows. WORLDCLIM data has 19 bioclimatic variables (https://www.worldclim.org/data/bioclim.html) (Amatulli et al., 2022), including 

* `BIO1` = Annual Mean Temperature
* `BIO2` = Mean Diurnal Range (Mean of monthly (max temp - min temp))
* `BIO3` = Isothermality (BIO2/BIO7) (×100)
* `BIO4` = Temperature Seasonality (standard deviation ×100)
* `BIO5` = Max Temperature of Warmest Month
* `BIO6` = Min Temperature of Coldest Month
* `BIO7` = Temperature Annual Range (BIO5-BIO6)
* `BIO8` = Mean Temperature of Wettest Quarter
* `BIO9` = Mean Temperature of Driest Quarter
* `BIO10` = Mean Temperature of Warmest Quarter
* `BIO11` = Mean Temperature of Coldest Quarter
* `BIO12` = Annual Precipitation
* `BIO13` = Precipitation of Wettest Month
* `BIO14` = Precipitation of Driest Month
* `BIO15` = Precipitation Seasonality (Coefficient of Variation)
* `BIO16` = Precipitation of Wettest Quarter
* `BIO17` = Precipitation of Driest Quarter
* `BIO18` = Precipitation of Warmest Quarter
* `BIO19` = Precipitation of Coldest Quarter

```{r environmental parameters from WORLDCLIM}

#Get climatic variables from the package folder

worldclim <- system.file('extdata/worldclim.tiff', package = 'specleanr')

worldclim <- terra::rast(worldclim)


```

2. **Preliminary analysis**

The step involves removing missing coordinate values, setting the geographical range for the data collated in **1.a**, if the user did not set the bounding box during data download. After that, the environmental data predictors are extracted. The user can set either parameter **list = TRUE** to return a list of datasets for each species, if multiple species are considered, or **list = FALSE** to return a combined dataframe. The Danube basin was collated from the Hydrography90m basin files (https://hydrography.org/hydrography90m/hydrography90m_layers).

```{r extract data and prelimianry analysis}

danube_basin <- sf::st_read(system.file('extdata', "danube.shp.zip", package = 'specleanr'), 
                            quiet = TRUE)

#Environmental predictors extraction for multiple species (multiple = TRUE)

plantcleaned <-  pred_extract(data= plantdf, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'species',  
                             bbox = danube_basin,
                             list= TRUE, 
                             minpts = 10, merge = FALSE, verbose = FALSE, warn = FALSE)

#Environmental prediction extraction for a single species (multiple = FALSE)
sp1df <- subset(plantdf, species=="Fagus sylvatica")

p1cleaned <-  pred_extract(data= sp1df, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'species',  
                             bbox = danube_basin, 
                             minpts = 10, merge = FALSE, 
                           verbose = FALSE, warn = FALSE)

```

3. **Detecting outliers using multiple outlier detection methods**
* To detect outliers we selected ensembled seven outlier detection methods at least from different categories i.e., 1) **univariate methods**: adjusted boxplot **`adjbox()`**, Hampel filter: **`hampel()`**, Z-score **`zscore()`** and reverse jackknifing **`jknife()`**; 2) **Multivariate methods or machine learning models**: local outlier factor: **`lof()`**, K-means: **`kmeans()`**, and Mahalanobis distance measure **`mahal()`**.

```{r oultlier detection}

#Flag outlier in single species data (multiple = TRUE)

outliersdata <- multidetect(data = plantcleaned,
                       multiple = TRUE,
                       var = 'bio1',
                       output = 'outlier',
                       exclude = c('x','y'), 
                       methods = c('adjbox', "hampel", 'zscore', 
                                   'lof', 'jknife', 'kmeans', 'mahal'),
                       silence_true_errors = FALSE, warn = F, verbose = F)

#Flag outlier in single species data (multiple = FALSE)

out1df <- multidetect(data = p1cleaned,
                       multiple = FALSE,
                       var = 'bio1',
                       output = 'outlier',
                       exclude = c('x','y'), 
                       methods = c('adjbox', "hampel", 'zscore', 
                                   'lof', 'jknife', 'kmeans', 'mahal'),
                       silence_true_errors = FALSE, warn = F, verbose = F)

```
**Visualizing outliers flagged by each method**

The plots are wrapped on a ggplot2 object and extensible with ggplot2 parameters. The **`ggoultiers()`** takes in 3 parameters: **1: x** the outlier object of **`datacleaner class`**; **2: y** the index or species name if the outlier has more than one species, and **3: raw** to either indicate the number of outlier if **TRUE** or proportional or outliers flagged about total number of occurrences if **FALSE**

```{r visualize outliers, fig.width = 6, fig.height= 3.5, fig.align='center'}

ggoutliers(x=outliersdata)

#for one species: no index needed
ggoutliers(x= out1df)

```
**Identify the best threshold using loess model.**

```{r threshold identifcation, fig.width = 6, fig.height= 4, fig.align='center'}

optimal1<- optimal_threshold(refdata = p1cleaned, outliers = out1df, plot = TRUE)


optimalthreshold <- optimal_threshold(refdata = plantcleaned, outliers = outliersdata, plot = FALSE)
```


*Indicate the best methods obtained for each species 

4. **Extracting clean data set after outlier detection**

The mode can be set to **mode = abs** to discard only absolute outliers or **mode = best** to use the best method. The outlier flagged by the best method will be used to clean the data. For more information, check (B....2024).

```{r extract outliers from clean dataset}

cleandf <- extract_clean_data(refdata = plantcleaned, outliers = outliersdata, 
                              threshold = 0.4, pabs = 0.1)


cleandf1 <- extract_clean_data(refdata = p1cleaned, outliers = out1df, 
                          threshold = 0.4, pabs = 0.1)

```

### References 

1. Amatulli, Giuseppe, et al. "Hydrography90m: A new high-resolution global hydrographic dataset." Earth System Science Data 14.10 (2022): 4525-4550.

2. Harvey-Brown, Y., Barstow, M., Mark, J. & Rivers, M.C. 2017. Populus nigra. The IUCN Red List of Threatened Species 2017: e.T63530A68106816. https://dx.doi.org/10.2305/IUCN.UK.2017-3.RLTS.T63530A68106816.en. Accessed on 24 June 2024.

3. de Rigo, D., Enescu, C.M., Houston Durrant, T. and Caudullo, G. 2016. Populus nigra in Europe: distribution, habitat, usage and threats. European Atlas of Forest Tree Species, Publications Office of the European Union, Luxembourg.

4. Barstow, M. & Beech, E. 2018. Fagus sylvatica. The IUCN Red List of Threatened Species 2018: e.T62004722A62004725. https://dx.doi.org/10.2305/IUCN.UK.2018-1.RLTS.T62004722A62004725.en. Accessed on 24 June 2024.





