---
title: "Detecting environmental outliers in environmental data for plants."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignettes for plants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE}
library(specleanr)
```

**Change in the distribution of *Salix alba*, *Fagus sylvatica*, and *Populus nigra* due to climate change**.

The three different species were randomly selected. The species *Populus nigra* is tree species under the Salicaceae family, which has been assessed 
has as data deficient by the International Union for Conservation of Nature. The tree species is native to Europe and predominantly in flood plain in 
mixed forests (IUCN 2013). The *Fagus sylvatica* 




```{r get species occurences}

plantdf <- getdata(data = c( "Salix alba", "Populus nigra", "Fagus sylvatica"), 
                    gbiflim = 700, inatlim = 100, isFish = FALSE,
                     hasCoordinate = TRUE, 
                   bbox = c(xmin = 8.15250, ymin = 42.08333, xmax=29.73583, ymax = 50.24500),
                   verbose = T, warn = T)

extrdata <- extract_online(plantdf)

```


```{r environmental parameters from WORLDCLIM}

#Get climatic variables from the package folder

worldclim <- system.file('extdata/worldclim.tiff', package = 'specleanr')

worldclim <- terra::rast(worldclim)

```

```{r}

danube <- system.file('extdata/danube/basinfinal.shp', package = 'specleanr')

danube_basin  <-     sf::st_read(danube, quiet=TRUE)


plantcleaned <-  pred_extract(data= extrdata, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'species',  
                             multiple = TRUE, 
                             bbox = danube_basin,
                             list= FALSE, 
                             minpts = 10, merge = FALSE, verbose = TRUE, warn = T)

xy <- plantcleaned %>% dplyr::filter(species=="Fagus sylvatica") #%>% dplyr::select(1:19)

#xf= xkmeans(data = xy, k =10, exclude = c("x", "y"), output = 'outlier', mode = 'soft', method = "silhouette")

plantlist <-  pred_extract(data= extrdata, 
                             raster= worldclim, 
                             lat = 'decimalLatitude',
                             lon = 'decimalLongitude',
                             colsp = 'species',  
                             multiple = TRUE, 
                             bbox = danube_basin,
                             list= TRUE, 
                             minpts = 10, merge = FALSE, verbose = TRUE, warn = T)
```


```{r}


plantoutliers1 <- multidetect(data = plantcleaned,
                       multiple = TRUE,
                       var = 'bio1',
                       colsp = "species",
                       output = 'outlier',
                       exclude = c('x','y'), 
                       methods = c('adjbox', "hampel", 'zscore', 'lof', 'jknife'),#,'iforest'
                       showErrors  = TRUE, warn = F, verbose = F)


p1<- multidetect(data = xy,
                       multiple = FALSE,
                       var = 'bio1',
                       output = 'outlier',
                       exclude = c('x','y'), 
                       methods = c('adjbox', "hampel", 'zscore', 'lof', 'jknife'),#,'iforest'
                       showErrors  = TRUE, warn = F, verbose = F)
```


```{r extract outliers from clean dataset}

x1 <- bestmethod(x=p1, threshold = 0.6, warn = TRUE, verbose = FALSE)

x2 <- bestmethod(x=plantoutliers1, threshold = 0.6, warn = TRUE, verbose = FALSE, sp="Fagus sylvatica")


bst <- multibestmethod(x=plantoutliers1, autothreshold = TRUE, warn = TRUE, verbose = FALSE)

```

##Data extraction
```{r}
# 
# c11 <- extract_clean_data(refdata = plantlist, outliers = plantoutliers1,
#                                 mode = "best",  threshold = 0.6, sp = "Fagus sylvatica",#
#                                 colsp = 'species', warn = T, verbose = T, pabs = 0.1)


# c2 <- extract_clean_data(refdata = xy, outliers = p1, mode = "best", threshold = 0.6, warn = T,
#                          verbose = TRUE, pabs = 0.1)

c2 <- multextract_clean(refdata = plantcleaned, outliers = plantoutliers1, 
                         mode = "abs", threshold = 1, warn = T, colsp = 'species',
                         verbose = TRUE, pabs = 0.1)


```

###Model comparison

```{r}

modecom <- modelcomparison(refdata = xy, outliers = plantoutliers1, raster = worldclim,
                           lat = 'y', lon = 'x', models = c("GLM"), mode = 'abs',  
                           thresholds = seq(0.2, 0.7, 0.1),
                           colsp = 'species', species = c("Fagus sylvatica"), full = FALSE)

#perf <- extract_performance(modecom)
```


```{r}
library(ggplot2)

ggplot(perf, aes(x=scenario, y=Specificity))+
  geom_boxplot()
```














